---
title: "Getting started with superassp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with superassp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(superassp)
have_audio <- requireNamespace("superassp", quietly = TRUE) &&
  nzchar(system.file("samples", "sustained", "a1.wav", package = "superassp"))
```

`superassp` is a self-contained R package for speech signal processing. It
bundles a large collection of DSP routines (pitch, formants, voice quality,
prosody, …) behind one consistent, `wrassp`-like interface, and produces output
that plugs directly into `emuR`. No external Praat or `wrassp` installation is
required.

## The two function families

Almost everything you call falls into one of two prefixes:

| Prefix   | Produces                                  | Output class   | Example                     |
|----------|-------------------------------------------|----------------|-----------------------------|
| `trk_*`  | A **time-series track** following the audio (F0, formants, RMS, …) | `AsspDataObj`  | `trk_pitch_rapt()`          |
| `lst_*`  | **Summary statistics** (scalars/vectors: jitter, voice-quality scores, …) | `JsonTrackObj` | `lst_voice_report()`        |

Two more families round things out: `ucnv_*` for unit conversions
(Hz↔Bark/Mel/ERB/semitone, dB↔phon/sone) and `read_*`/`write_*` for I/O
(`read_audio`, `read_ssff`, `read_jstf`, and their `write_` counterparts).

## Output modes: in memory vs. to file

Every `trk_*` and `lst_*` function takes a `toFile` argument:

* `toFile = FALSE` — return the result **in memory** (an `AsspDataObj` or
  `JsonTrackObj`) for immediate use in R. Single file only.
* `toFile = TRUE` (the default for most functions) — **write** an SSFF track
  file (`trk_*`) or JSTF file (`lst_*`) next to each input, and return a count.
  This is the batch mode; pass a vector of paths and it parallelises across files.

```{r, eval = have_audio}
wav <- system.file("samples", "sustained", "a1.wav", package = "superassp")

# In-memory analysis
rms <- trk_rms(wav, toFile = FALSE, verbose = FALSE)
rms
```

## Inspecting an `AsspDataObj`

Use the accessor generics rather than digging into attributes:

```{r, eval = have_audio}
track_names(rms)       # which tracks are present
sample_rate(rms)       # frame rate in Hz
n_records(rms)         # number of analysis frames
signal_duration(rms)   # seconds
```

`as.data.frame()` (or `as_tibble()`) flattens all tracks into one time-indexed
table — the usual bridge to `dplyr` and plotting:

```{r, eval = have_audio}
df <- as.data.frame(rms)
head(df)
```

Individual track matrices are reached by name, e.g. `rms[["RMS[dB]"]]`.

## Inspecting a `JsonTrackObj`

Summary (`lst_*`) functions return a `JsonTrackObj`: a self-describing container
of measures with their schema and provenance.

```{r, eval = FALSE}
vq <- lst_voice_report(wav, toFile = FALSE)
vq                       # compact summary
names(vq$field_schema)   # available measures
vq$slices[[1]]$values    # values for the first slice
```

## Which pitch tracker should I use?

All pitch trackers emit a fundamental-frequency track (0 marks unvoiced frames).
They differ in method, speed, and robustness:

| Function              | Method / origin                      | Notes                                         |
|-----------------------|--------------------------------------|-----------------------------------------------|
| `trk_pitch_rapt`      | RAPT autocorrelation (SPTK)          | Fast, robust general-purpose default          |
| `trk_pitch_swipe`     | SWIPE′ (SPTK)                        | Accurate, good on noisy speech                |
| `trk_pitch_reaper`    | REAPER (Google)                      | Also yields voicing / GCI information         |
| `trk_pitch_yin`       | YIN                                  | Classic difference-function tracker; +probability |
| `trk_pitch_pyin`      | probabilistic YIN                    | Smoother voicing decisions than YIN           |
| `trk_pitch_crepe`     | CREPE deep neural net (ONNX)         | Most accurate; heavier (downloads a model)    |
| `trk_pitch_srh`       | Summation of Residual Harmonics      | Strong under additive noise                   |
| `trk_pitch_dio` / `trk_pitch_harvest` | DIO / Harvest (WORLD)| Fast (DIO) vs. accurate (Harvest)             |
| `trk_pitch_ac` / `trk_pitch_cc` / `trk_pitch_shs` / `trk_pitch_spinet` | Praat (autocorrelation / cross-correlation / subharmonic / SPINET) | Praat-faithful behaviour |
| `trk_pitch_ksv` / `trk_pitch_mhs` | ASSP (K. Schäfer-Vincent / Michel) | Bundled C library, no external deps  |

Start with `trk_pitch_rapt` (fast, reliable). Reach for `trk_pitch_swipe` or
`trk_pitch_crepe` when you need maximum accuracy, and `trk_pitch_srh` for noisy
recordings.

## Which formant tracker should I use?

| Function                   | Method / origin                    | Notes                                    |
|----------------------------|------------------------------------|------------------------------------------|
| `trk_formant_burg`         | Praat Burg LPC (via pladdrr)       | Praat-faithful default                   |
| `trk_formant_forest`       | ASSP forest tracker                | Bundled C, no external deps              |
| `trk_formant_snack`        | Snack ESPS-style LPC               | Familiar to Snack/WaveSurfer users       |
| `trk_formant_cgdzp`        | Complex-group-delay (COVAREP)      | Robust bandwidth estimates               |
| `trk_formant_tvwlp`        | Time-varying weighted LP           | Better on rapid transitions              |
| `trk_formant_deepformants` / `trk_formant_formantnet` | Deep neural nets (ONNX) | Highest accuracy; heavier |

Start with `trk_formant_burg` or `trk_formant_forest`. Use the neural trackers
(`trk_formant_deepformants`, `trk_formant_formantnet`) when accuracy matters more
than runtime.

## Batch processing to files

```{r, eval = FALSE}
files <- c("speaker1.wav", "speaker2.wav", "speaker3.wav")

# Writes one .rap SSFF file next to each input, parallelised across files
trk_pitch_rapt(files, toFile = TRUE)
```

The written SSFF/JSTF files load straight back with `read_ssff()` / `read_jstf()`
and are ready for use with `emuR`.
