---
title: "SSFF and JSTF: reading, writing, and round-tripping"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SSFF and JSTF: reading, writing, and round-tripping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(superassp)
have_audio <- requireNamespace("superassp", quietly = TRUE) &&
  nzchar(system.file("samples", "sustained", "a1.wav", package = "superassp"))
# lst_voice_report() is backed by the optional, GitHub-only 'pladdrr' package.
# Chunks that call it are skipped when it is absent, so the vignette still
# builds on machines (CRAN, win-builder) that do not have pladdrr installed.
have_pladdrr <- requireNamespace("pladdrr", quietly = TRUE)
```

`trk_*` functions write **SSFF** (Simple Signal File Format) tracks;
`lst_*` functions write **JSTF** (JSON Track Format) summaries. Both
round-trip through matching `read_*`/`write_*` pairs and both load
straight into `emuR`.

## SSFF: time-series tracks

Write a track to disk, then read it back with `read_ssff()`:

```{r, eval = have_audio}
wav <- system.file("samples", "sustained", "a1.wav", package = "superassp")
out_dir <- tempdir()

trk_pitch_rapt(wav, toFile = TRUE, outputDirectory = out_dir, explicitExt = "f0")

f0_path <- file.path(out_dir, paste0(tools::file_path_sans_ext(basename(wav)), ".f0"))
f0_obj <- read_ssff(f0_path)
track_names(f0_obj)
sample_rate(f0_obj)
```

`write_ssff()` is the inverse operation, for writing an `AsspDataObj` you
built or modified in R:

```{r, eval = have_audio}
f0_obj[["f0"]][1:5, ] <- 0  # zero out the first 5 frames
out_path <- tempfile(fileext = ".f0")
write_ssff(f0_obj, out_path)
read_ssff(out_path)[["f0"]][1:5, ]
```

`read_track()`/`write_track()` are format-agnostic dispatchers: pass any
`trk_*` output path and they detect SSFF vs. JSTF from the file's own
metadata rather than the file extension.

## JSTF: summary measures

`lst_*` functions produce a `JsonTrackObj` — a self-describing container
with a field schema and one "slice" per analysis window:

```{r, eval = have_audio && have_pladdrr}
vr <- lst_voice_report(wav, toFile = FALSE, return_jstf = TRUE)
names(vr$field_schema)
vr$slices[[1]]$values
```

Write it to a `.jstf` file and read it back:

```{r, eval = have_audio && have_pladdrr}
jstf_path <- tempfile(fileext = ".jstf")
write_jstf(vr, jstf_path)
vr_reloaded <- read_jstf(jstf_path)
identical(vr$field_schema, vr_reloaded$field_schema)
```

## Building a `JsonTrackObj` from your own results

If you're wrapping a measure that isn't already a `lst_*` function,
`create_json_track_obj()` builds the same self-describing structure that
`write_jstf()` expects. It is an internal helper (not exported — see the
package's Export Policy), so package contributors writing a new `lst_*`
wrapper call it via `:::`:

```{r, eval = have_audio}
obj <- superassp:::create_json_track_obj(
  results = list(f0_mean = 150.2, f0_sd = 18.4),
  function_name = "my_custom_summary",
  file_path = wav,
  sample_rate = 44100,
  audio_duration = 1.5
)
write_jstf(obj, tempfile(fileext = ".jstf"))
```

## Loading into emuR

Both `read_ssff()`-produced `AsspDataObj`s and `read_jstf()`-produced
`JsonTrackObj`s carry the sample rate, start time, and track names emuR
expects — write with `toFile = TRUE` into your emuR database's `_ssff`
directory structure and the tracks load without further conversion.
