---
title: "Voice quality and creak"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Voice quality and creak}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(superassp)
have_audio <- requireNamespace("superassp", quietly = TRUE) &&
  nzchar(system.file("samples", "sustained", "a1.wav", package = "superassp"))
have_pladdrr <- requireNamespace("pladdrr", quietly = TRUE)
```

Voice quality analysis in `superassp` spans three layers: glottal-source
extraction (`trk_*` functions returning a per-sample or per-frame track),
creak/GCI detection (`trk_*` functions returning event tracks), and
whole-recording summary scores (`lst_*` functions).

## Glottal flow: Iterative Adaptive Inverse Filtering

`trk_covarep_iaif()` separates the glottal source from the vocal tract,
returning the glottal flow and its derivative (MFDR) at the native audio
sample rate — the standard input to jitter/shimmer/HNR-style measures.

```{r, eval = have_audio}
wav <- system.file("samples", "sustained", "a1.wav", package = "superassp")

iaif <- trk_covarep_iaif(wav, toFile = FALSE, verbose = FALSE)
track_names(iaif)
```

## Detecting creaky voice

Two creak detectors are available with the same output schema
(`creak_pp`: posterior probability; `creak_bin`: binary decision), so they
are interchangeable in downstream code:

* `trk_creak_vat()` — a bit-faithful Rcpp port of the Kane-Drugman + Ishi
  ANN pipeline from the MATLAB Voice Analysis Toolkit. Prefer this when
  parity with the original MATLAB VAT output matters.
* `trk_covarep_creak()` — a partial R-side reimplementation; may differ
  from `trk_creak_vat()` on borderline frames.

```{r, eval = have_audio}
creak <- trk_creak_vat(wav, toFile = FALSE)
head(as.data.frame(creak))
```

Glottal closure instants (GCIs) — the time points used to anchor many
voice-quality measures — are available separately via `trk_gci_vat()`.

## Whole-recording voice quality summaries

`lst_vq()` runs the Praat-based `VQ_measurements_V2.praat` pipeline
(two-pass adaptive pitch detection followed by 36 voice-quality
parameters: jitter, shimmer, multi-band HNR, spectral energy measures,
and glottal-to-noise excitation ratio) and returns one row per file.

```{r, eval = have_audio && have_pladdrr}
vq <- lst_vq(wav, toFile = FALSE, verbose = FALSE)
str(vq)
```

Restrict the initial pitch search range for atypical voices (e.g. children,
high-pitched pathological voices) via `minPitchInitial`/`maxPitchInitial`:

```{r, eval = FALSE}
lst_vq("child.wav", minPitchInitial = 150, maxPitchInitial = 600)
```

## Which function should I use?

| Goal                                   | Function              | Requires   |
|-----------------------------------------|-----------------------|------------|
| Glottal flow / MFDR at sample rate      | `trk_covarep_iaif()`  | —          |
| Creak probability, MATLAB-VAT parity    | `trk_creak_vat()`     | —          |
| Creak probability, lighter R-side path  | `trk_covarep_creak()` | —          |
| Glottal closure instants                | `trk_gci_vat()`       | —          |
| 36-parameter voice-quality summary      | `lst_vq()`            | `pladdrr`  |
